---
title: "Exploring the BRFSS data"
Author: " Rudy Meza"
Date : "2020-08-24"
output: github_document

---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data



```{r load-data}
load("C:/Users/rudym/OneDrive/Desktop/Data Analysis/Epidemic of Obesity/brfss2013.rdata")
```
```{r}
data<-brfss2013%>%select(genhlth,cvdcrhd4,X_bmi5,X_bmi5cat,bphigh4,toldhi2,cvdstrk3,educa,strength,smoke100,alcday5,exerany2,X_ageg5yr,exeroft2,imonth,iday, iyear, X_age_g, smokday2)

str(data)
na_vec<-which(!complete.cases(data))
data<-data[-na_vec,]
str(data)
attach(data)
```



* * *

## Part 1: Data
The Behavioral Risk Factor Surveillance System (BRFSS) data is a telephone survey in the U.S. and all U.S. territories. The data collection method is uniform and random. The goal of this survey is to collect data on preventative health factors and risky behavior that have been linked to injuries, chronic illness, and disease. This would be a generalized analysis to link many factors that affect a person's health. This is an observational study conducted by the BRFSS.



* * *

## Part 2: Research questions
Authors Hammond and Levine analyze the types of economic costs associated with obesity. They describe 4 types of costs that impact the economy linked to obesity. These are medical, transportation, productivity, and human capital. They specifically link five types of diseases to obesity including stroke, hypertension, hypercholesterolemia, coronary heart disease, and diabetes. I want to analyze these five links listed in their' analyses. 

**Research quesion 1:**
Given the listed diseases associated with obesity, which disease has the highest correlations with obesity?

**Research quesion 2:**
What is the average of obesity per month? How does month to month rates compare?

**Research quesion 3:**
How do individuals who exercise compare in the BMI category? 
What about individuals who smoke and are in the obese category?
* * *



## Part 3: Exploratory data analysis



**Research quesion 1:**


```{r}
 high<- data %>% 
 select(X_bmi5cat,cvdcrhd4,cvdstrk3,bphigh4,toldhi2, iyear, imonth)
```
```{r}
high<-high %>% 
 mutate(obese=ifelse(X_bmi5cat=="Obese",1,0)) %>%
  mutate(heart_disease=ifelse(cvdcrhd4=="Yes",1,0)) %>%
  mutate(stroke=ifelse(cvdstrk3=="Yes",1,0))  %>%
  mutate(bp=ifelse(bphigh4=="Yes",1,0)) %>%
  mutate(cho=ifelse(toldhi2=="Yes",1,0))
```
```{r}
corr<-round(cor(high[,8:12]),2)
corr
```
Here we can see that individuals with High Blood Pressure have the highest correlation with obesity. High Cholesterol has the highest correlation with high blood pressure.

```{r}
c_plot <-high %>%
  filter(bphigh4=="Yes") %>%
  group_by(X_bmi5cat) 
  
  
ggplot(c_plot, aes(x=X_bmi5cat, fill=X_bmi5cat))+
  geom_bar()+ggtitle("BMI CATAGORY per HIGH BLOOD PRESSURE")+theme(plot.title=element_text(hjust=0.5))
```



We can see that there are more who are overweight compared to obese that have been diagnosed with high blood pressure.


**Research quesion 2:**
```{r}
high_ob<-high %>%
  select(obese, imonth) %>%
  group_by(imonth) %>%
  summarise(ob_rate=sum(obese)/n()) 
            

  ggplot(data=high_ob, aes(x=as.numeric(imonth), y=ob_rate))+
  geom_point() +
    geom_line()
```

Here we can see there are spikes during the summertime. There is a turn around in the summer followed by an increase than a decrease in the fall. One last spike around November and then a decrease in December.

**Research quesion 3:**


```{r}
high<- high %>%
  mutate(exer=ifelse(exerany2=="Yes",1,0))
```
```{r}
high_ex<-data %>%
  select(exerany2,X_bmi5cat) %>%
  group_by(X_bmi5cat,exerany2) %>%
  summarise(count=n())
```
```{r}
ggplot(data=high_ex)+ 
  geom_bar(mapping=aes(x=X_bmi5cat,y=count, fill=X_bmi5cat), stat="identity")
```

Here we see that individuals who are obese do participate in walking, jogging, and running outside of work. We also see that underweight individuals have the lowest count for physical activity outside of work.

```{r}
smoke_d<- data %>%
 select(X_bmi5cat,exeroft2,smokday2,X_age_g)%>%
  filter(X_bmi5cat=="Obese", smokday2=="Every day",exeroft2<=210)%>%
  group_by(X_age_g)

```

```{r}
  
ggplot(data=smoke_d, aes(x=factor(X_age_g), y=exeroft2, fill=X_age_g))+
 geom_boxplot()
```

These plots show that the age range for walking, jogging, and running have a median close to 110 times per week. These are individuals who smoke every day and are considered obese.